BlueTracer: \\ a Robust API Tracer for Evasive Malware

### 2018-07-02 implementation.tex
Blue Pill Added

We will start the current chapter by giving a general overview of BlueTracer. Specifically we will describe the main characteristics of Intel Pin, the DBI framework on which BlueTracer is based on. Then we will introduce Blue Pill, the DBI-based software toolkit allowing BlueTracer to be robust against anti-evasive malware. Lastly, we will illustrate BlueTracer's structure, which will serve as a baseline for the rest of the chapter.

@@

\subsection{Blue Pill}
\subsection{Structure}

BlueTracer has been organized primarily taking into account the rich set of APIs offered by Pin, which have led to the decision to split the tool in three parts: the first aimed at \textbf{native APIs tracing}, the second for \textbf{APIs tracing} and the last focused on \textbf{callbacks and APCs tracing} (\textit{Figure 3.1}). In particular, the tracing of native APIs also employs a different source of API information (\textit{Dr. Memory}'s system call data) than the ones utilized for tracing APIs (\textit{drltrace}'s configuration file or the information extracted from \textit{PyREbox}'s database).

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.85\textwidth]{Figures/BlueTracer.pdf}
\caption{\textit{BlueTracer's high level structure}}
\end{figure}

@@

In BlueTracer, the traced data is written to a binary file, one for every thread. The default naming convention used for the tracer's log files is \texttt{Traced.[OS Process ID].[Pin-specific Thread ID]}, similarly to the one Blue Pill employs in its own log files, with the user being able to change \texttt{Traced} with a name of its choice in the configuration file.

@@

Pin provides a set of APIs aimed at assisting the extraction of information relative to the system calls made by the pinned application, also including information relative to Windows Native APIs. In particular, in Blue Pill (and consequently in BlueTracer) \texttt{PIN\_AddSyscallEntryFunction} and  \texttt{PIN\_AddSyscallExitFunction} were used for this purpose. In fact, these allow to register notification functions which are called immediately before and after the execution of a system call \cite{Pin}. In BlueTracer, the function in charge of gathering Native API information before execution is \texttt{TraceSysCallEntry} while the one responsible for collecting Native API information after execution is \texttt{TraceSysCallExit}. They have been both been embedded in Blue Pill's notification functions and their overall structure is detailed in \textit{Figure 3.2}.
\\
\begin{figure}[h]
\centering
\includegraphics[scale=0.75]{Figures/NTAPITracing.pdf}
\caption{\textit{Native APIs Tracing Workflow}}
\end{figure}
\\
Following \textit{Figure 3.2}, let us now thoroughly analyze the main steps which take place when Native APIs are traced.

\subsection{Main Image Check}
In order to filter the logged information relative to Native APIs, BlueTracer allows the analyst to decide, through the use of a boolean parameter (\texttt{MainImage}) in the configuration file, whether or not only Native APIs called invoked directly from the main executable of the pinned application should be traced. Therefore, as it can be seen in \textit{Figure 3.2}, the first thing which is done in \texttt{TraceSysCallEntry}, assuming that the \texttt{MainImage} configuration parameter has been set to \texttt{true}, is to determine if the Native API call is taking place directly from the main executable.

To this end, Pin's \texttt{IMG} APIs are employed, where in Pin an \texttt{IMG} represents all the data structures relative to binaries and shared libraries \cite{Pin}. Specifically, in Blue Pill, \texttt{IMG\_AddInstrumentFunction} is utilized to register a callback which is invoked each time an image is loaded. Inside such callback, which takes as one of the input parameters the \texttt{IMG} object representing the image being loaded, the following steps are taken:
\begin{enumerate}
\item \texttt{IMG\_IsMainExecutable} is employed to determine if the image being loaded is the main executable of the pinned application.
\item If this is the case, \texttt{IMG\_HighAddress} and \texttt{IMG\_LowAddress} are invoked. These two APIs return the highest address and the lowest address respectively of any code or data loaded by the image corresponding the the \texttt{IMG} object they take as input. By employing these APIs is therefore possible to determine the address range relative to the main executable. 
\end{enumerate}

@@

In Pin, the Native API is intercepted right before \texttt{sysenter} is executed. This could be inferred by the fact that, in \texttt{TraceSysCallEntry}, the instruction pointer (obtainable via the \texttt{PIN\_GetContextReg} API) contains \texttt{sysenter}'s address. In light of this, in addition to the fact that each Native API calls \texttt{KiFastSystemCall} without setting a stack frame, the application's stack during the execution of \texttt{TraceSysCallEntry} is in the following state (\textit{Figure 3.3}).
\\
\begin{figure}[h]
\centering
\includegraphics[width=0.7\textwidth]{Figures/StackMainImage.pdf}
\caption{\textit{Stack before} \texttt{sysenter} \textit{execution}}
\end{figure}

@@

\begin{enumerate}
\item The first check involves discovering whether or not \texttt{sysenter} is called directly from the main executable. As Pin intercepts Native APIs right before the execution of \texttt{sysenter}, this can be trivially done by checking if the instruction pointer value obtained in \texttt{TraceSysCallEntry} falls within the main executable memory address range.
\item Secondly, it is required to examine if the Native API return address belongs to the main executable. As a result of what shown in \textit{Figure 3.3} such return address is obtained by adding 4 to the stack pointer and retrieving the pointed value, where the stack pointer can be obtained in Pin through the use of \texttt{PIN\_GetContextReg}. The resulting value is then, once more, compared with the main executable memory address range.
\end{enumerate}

@@

Previously, it was stated how it is possible to register a callback function to be executed before Native APIs through the use of \texttt{PIN\_AddSyscallEntryFunction}. Such callback functions, named \texttt{SYSCALL\_ENTRY\_CALLBACK}s, receive a set of parameters, including \texttt{ctx}, the application's register state immediately before the system call execution, and \texttt{std}, the system call standard. In Blue Pill, these two parameters are used to invoke \texttt{PIN\_GetSyscallNumber} inside its \texttt{SYSCALL\_ENTRY\_CALLBACK}. Such API returns the number (ID) of the Native API to be executed in the provided context \cite{Pin}.

Blue Pill employs a mechanism which allows it to obtain the Native API name from its number, since the hooking functions it employs are indexed by name and not by ordinal. This choice was made because the identifiers vary depending on the Windows version, even among different Service Pack versions. For the same reason, the information needed by BlueTracer to correctly log Native APIs is also indexed by Native API name and, consequently, BlueTracer also utilizes the aforementioned Native API name resolution mechanism.

@@

In \textit{3.2.2} it was explained how BlueTracer obtains a Native API name from its ID. After that, as shown in \textit{Figure 3.2}, there must be a way to, given a specific Native API name, retrieve the corresponding \texttt{syscall\_info\_t struct}. This is done by building an hash map during the tool's initialization which maps the Native API name to the appropriate \texttt{struct}. Such idea is also adopted for named constants enumerations since a hash map is also constructed with the objective of associating the name of a named constant enumeration to the corresponding \texttt{struct}.

\subsection{Native API Logging}
With reference to \textit{Figure 3.2}, let us consider the point in \texttt{TraceSysCallEntry} where the \texttt{syscall\_info\_t struct} relative to the Native API being traced was successfully retrieved. Before describing how the logging of "known" Native APIs is performed, it is worth mentioning that, in case a \texttt{struct} describing a certain Native API is not found, BlueTracer just logs the name of the Native API and the input value of a user-specified number of arguments (4 by default).

@@

To conclude this section, let us show how a traced Native API looks like in the log file (\textit{Figure 3.4}). As it can be observed, the adopted log format is quite intuitive. In fact, the idea was to have a log which is both easy to read by the analyst but also easy to parse and post-process.      
\\

@@

When tracing APIs, most of the concepts seen in Native API tracing also apply, but there are also significant differences, both in the steps that were carried out and in the Pin APIs which were employed, as it can be observed from the diagram in \textit{Figure 3.4}.

Now let us illustrate the actions being performed during API tracing by analyzing in detail each one of the three main blocks of \textit{Figure 3.4}.

@@

The first step taken towards APIs tracing is APIs instrumentation, i.e. placing analysis routines before and after the APIs' execution. To do this, a specific function was embedded in Blue Pill's image callback (the notion of image callback was previously introduced in Section 3.3.1). Such function, named \texttt{AddAnalysisRoutines}, performs the following actions for every DLL being loaded:

@@

Due to the high amount of APIs for which information was stored in the \texttt{PyREbox} database, it was decided to group API data by DLL. Specifically, for every DLL a corresponding \texttt{.cpp} file was created to store the API-related information under the form of \texttt{libcall\_arg\_info\_t} instances, one for each "known" API exported by that DLL. Furthermore, in order to make API data retrieval more efficient, it was chosen to employ a two-level hash map approach for this purpose. Given the DLL name (easily obtainable in the image callback via the use of \texttt{IMG\_Name}), the first level hash map is utilized to retrieve the DLL-specific second level hash map, which, in turn, maps the API name to the appropriate \texttt{struct} instance (\textit{Figure 3.5}). This is how API information retrieval is carried out in case \textit{PyREbox} database is used as API data source.
\\
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/HashMap.pdf}
\caption{\textit{PyREbox API Information Retrieval}}
\end{figure}

@@

In this section we will describe how APIs are analyzed before execution, i.e. the main steps that are carried out in the \texttt{PrintInfoBefore} functions, an overview of which is given in \textit{Figure 3.4}.
The workflow in this case is mostly the same as the one for NTAPIs' analysis, with one major difference, that is the introduction of a shadow stack used to store API entries in the thread local storage, which will be motivated later in the section.

@@

Lastly, let us detail the most important operations performed towards tracing API-related information after the API execution. These are carried out in the \texttt{PrintInfoAfter} function and can be observed from \textit{Figure 3.4}.

@@

Right before the execution of \texttt{KiUserCallbackDispatcher}, i.e. when the registered \texttt{CONTEXT\_CALLBACK\_NOTIFICATION} function is called, the stack looks as follows (\textit{Figure 3.6}).
\\
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/StackCallback.pdf}
\caption{\textit{Stack before \texttt{KiUserCallbackDispatcher} execution}}
\end{figure}

@@

\begin{enumerate}
\item The stack pointer value is retrieved. This is achieved through the use of the \texttt{PIN\_GetContextReg} API, that allows to get the value of a register in the context provided as input. Such context, in our specific case, is the application's register state after the context change, which, as previously mentioned, is passed as argument to the \texttt{CONTEXT\_CHANGE\_CALLBACK} notification function.
\item The callback number is obtained by adding 4 to the stack pointer and fetching the pointed value, in accordance to \textit{Figure 3.6}.
\item The \texttt{KernelCallbackTable} is retrieved from the PEB. 
\item The function pointer stored in the \texttt{KernelCallbackTable} entry indicated by the callback number is acquired.
\item It is checked if the address of the function falls within \texttt{User32.dll} memory address range. In fact, in \texttt{Win32} user mode callbacks are employed exclusively by \texttt{User32.dll} for windowing related aspects \cite{UCDispatcher}, meaning that the retrieved function address must belong to \texttt{User32.dll}'s memory address range. Such range is determined by retrieving an handle for \texttt{User32.dll} through \texttt{GetModuleHandle}, building an \texttt{IMG} object from it utilizing Pin's \texttt{IMG\_FindByAddress} and invoking the usual \texttt{IMG\_LowAddress} and \texttt{IMG\_HighAddress} functions. 
\item The name of the callback is determined by creating a \texttt{RTN} object from the function address via \texttt{RTN\_FindByAddress} and consequently invoking \texttt{RTN\_Name} on such object.
\item The callback occurrence and the callback name is recorded in the log.
\end{enumerate}

>>>>

We will start the current chapter by giving a general overview of BlueTracer. Specifically we will describe the main characteristics of Intel Pin, the DBI framework on which BlueTracer is based on. Then we will introduce BluePill, the DBI-based software toolkit allowing BlueTracer to be robust against anti-evasive malware. Lastly, we will illustrate BlueTracer's structure, which will serve as a baseline for the rest of the chapter.

@@

\subsection{BluePill}

BlueTracer remains undetected from evasive malware thanks to its integration with BluePill, which essentially constitutes BlueTracer's foundations. The main concept behind BluePill is the idea of using DBI to deceive highly evasive malware instances into believing they are executing in the environment they are designed to attack.

BluePill \cite{BluePill} is a software toolkit built on top of the Intel Pin DBI framework which:

\begin{itemize}
\item Allows the simulation of a real production environment a specific malware sample was intended for.
\item Conceals any virtualization artifacts and software setup which might set off evasion. 
\end{itemize}

In particular, BluePill implements over 100 Windows anti-evasion techniques as a set of low-level primitives aimed at monitoring, inspecting and faking the outcomes of malware operations investigating the execution environment. Experimental results have validated BluePill's effectiveness and have shown its ability to hinder a wide range of evasion mechanisms.

One of the BluePill's core components is the \textit{Target platform simulation layer}. It is in charge of managing how API calls, system calls and special instructions leak execution environment-specific information to the malware instance. In fact, details like operating system, hardware and the execution environment need to be tailored to meet the sample's expectations. BlueTracer operates at this level. In fact, BlueTracer complements BluePill by providing a detailed account of the malware instance activity, allowing the analysts to understand in detail what the malware's actions are. So, while BluePill's \textit{Target platform simulation layer} is aimed at deceiving the malware sample, BlueTracer tracks the information related to the malware activity before it is tricked, allowing to determine what the malicious sample was attempting to do. In this way it is possible to track in detail the activity of a malware sample, a feature BluePill lacks, while, at the same time, not worrying about the sample's anti-evasion activity, since this is addressed by BluePill.

In light of what just stated, BlueTracer's integration with BluePill is straightforward, since it essentially complements the \textit{Target platform simulation layer}, as it can be observed from \textit{Figure 3.1}.

\subsection{Structure}

BlueTracer has been organized primarily taking into account the rich set of APIs offered by Pin, which have led to the decision to split the tool in three parts: the first aimed at \textbf{native APIs tracing}, the second for \textbf{APIs tracing} and the last focused on \textbf{callbacks and APCs tracing} (\textit{Figure 3.2}). In particular, the tracing of native APIs also employs a different source of API information (\textit{Dr. Memory}'s system call data) than the ones utilized for tracing APIs (\textit{drltrace}'s configuration file or the information extracted from \textit{PyREbox}'s database).

@@

\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/BlueTracer.pdf}
\caption{\textit{BlueTracer's high level structure}}
\end{figure}

@@

In BlueTracer, the traced data is written to a binary file, one for every thread. The default naming convention used for the tracer's log files is \texttt{Traced.[OS Process ID].[Pin-specific Thread ID]}, similarly to the one BluePill employs in its own log files, with the user being able to change \texttt{Traced} with a name of its choice in the configuration file.

@@

Pin provides a set of APIs aimed at assisting the extraction of information relative to the system calls made by the pinned application, also including information relative to Windows Native APIs. In particular, in BluePill (and consequently in BlueTracer) \texttt{PIN\_AddSyscallEntryFunction} and  \texttt{PIN\_AddSyscallExitFunction} were used for this purpose. In fact, these allow to register notification functions which are called immediately before and after the execution of a system call \cite{Pin}. In BlueTracer, the function in charge of gathering Native API information before execution is \texttt{TraceSysCallEntry} while the one responsible for collecting Native API information after execution is \texttt{TraceSysCallExit}. They have been both been embedded in BluePill's notification functions and their overall structure is detailed in \textit{Figure 3.3}.
\\
\begin{figure}[h]
\centering
\includegraphics[scale=0.75]{Figures/NTAPITracing.pdf}
\caption{\textit{Native APIs Tracing Workflow}}
\end{figure}
\\
Following \textit{Figure 3.3}, let us now thoroughly analyze the main steps which take place when Native APIs are traced.

\subsection{Main Image Check}
In order to filter the logged information relative to Native APIs, BlueTracer allows the analyst to decide, through the use of a boolean parameter (\texttt{MainImage}) in the configuration file, whether or not only Native APIs called invoked directly from the main executable of the pinned application should be traced. Therefore, as it can be seen in \textit{Figure 3.3}, the first thing which is done in \texttt{TraceSysCallEntry}, assuming that the \texttt{MainImage} configuration parameter has been set to \texttt{true}, is to determine if the Native API call is taking place directly from the main executable.

To this end, Pin's \texttt{IMG} APIs are employed, where in Pin an \texttt{IMG} represents all the data structures relative to binaries and shared libraries \cite{Pin}. Specifically, in BluePill, \texttt{IMG\_AddInstrumentFunction} is utilized to register a callback which is invoked each time an image is loaded. Inside such callback, which takes as one of the input parameters the \texttt{IMG} object representing the image being loaded, the following steps are taken:
\begin{enumerate}
\item \texttt{IMG\_IsMainExecutable} is employed to determine if the image being loaded is the main executable of the pinned application.
\item If this is the case, \texttt{IMG\_HighAddress} and \texttt{IMG\_LowAddress} are invoked. These two APIs return the highest address and the lowest address respectively of any code or data loaded by the image corresponding the the \texttt{IMG} object they take as input. By employing these APIs is therefore possible to determine the address range relative to the main executable. 
\end{enumerate}

@@

In Pin, the Native API is intercepted right before \texttt{sysenter} is executed. This could be inferred by the fact that, in \texttt{TraceSysCallEntry}, the instruction pointer (obtainable via the \texttt{PIN\_GetContextReg} API) contains \texttt{sysenter}'s address. In light of this, in addition to the fact that each Native API calls \texttt{KiFastSystemCall} without setting a stack frame, the application's stack during the execution of \texttt{TraceSysCallEntry} is in the following state (\textit{Figure 3.4}).
\\
\begin{figure}[h]
\centering
\includegraphics[width=0.7\textwidth]{Figures/StackMainImage.pdf}
\caption{\textit{Stack before} \texttt{sysenter} \textit{execution}}
\end{figure}

@@

\begin{enumerate}
\item The first check involves discovering whether or not \texttt{sysenter} is called directly from the main executable. As Pin intercepts Native APIs right before the execution of \texttt{sysenter}, this can be trivially done by checking if the instruction pointer value obtained in \texttt{TraceSysCallEntry} falls within the main executable memory address range.
\item Secondly, it is required to examine if the Native API return address belongs to the main executable. As a result of what shown in \textit{Figure 3.4} such return address is obtained by adding 4 to the stack pointer and retrieving the pointed value, where the stack pointer can be obtained in Pin through the use of \texttt{PIN\_GetContextReg}. The resulting value is then, once more, compared with the main executable memory address range.
\end{enumerate}

@@

Previously, it was stated how it is possible to register a callback function to be executed before Native APIs through the use of \texttt{PIN\_AddSyscallEntryFunction}. Such callback functions, named \texttt{SYSCALL\_ENTRY\_CALLBACK}s, receive a set of parameters, including \texttt{ctx}, the application's register state immediately before the system call execution, and \texttt{std}, the system call standard. In BluePill, these two parameters are used to invoke \texttt{PIN\_GetSyscallNumber} inside its \texttt{SYSCALL\_ENTRY\_CALLBACK}. Such API returns the number (ID) of the Native API to be executed in the provided context \cite{Pin}.

BluePill employs a mechanism which allows it to obtain the Native API name from its number, since the hooking functions it employs are indexed by name and not by ordinal. This choice was made because the identifiers vary depending on the Windows version, even among different Service Pack versions. For the same reason, the information needed by BlueTracer to correctly log Native APIs is also indexed by Native API name and, consequently, BlueTracer also utilizes the aforementioned Native API name resolution mechanism.

@@

In \textit{3.2.2} it was explained how BlueTracer obtains a Native API name from its ID. After that, as shown in \textit{Figure 3.3}, there must be a way to, given a specific Native API name, retrieve the corresponding \texttt{syscall\_info\_t struct}. This is done by building an hash map during the tool's initialization which maps the Native API name to the appropriate \texttt{struct}. Such idea is also adopted for named constants enumerations since a hash map is also constructed with the objective of associating the name of a named constant enumeration to the corresponding \texttt{struct}.

\subsection{Native API Logging}
With reference to \textit{Figure 3.3}, let us consider the point in \texttt{TraceSysCallEntry} where the \texttt{syscall\_info\_t struct} relative to the Native API being traced was successfully retrieved. Before describing how the logging of "known" Native APIs is performed, it is worth mentioning that, in case a \texttt{struct} describing a certain Native API is not found, BlueTracer just logs the name of the Native API and the input value of a user-specified number of arguments (4 by default).

@@

To conclude this section, let us show how a traced Native API looks like in the log file (\textit{Figure 3.5}). As it can be observed, the adopted log format is quite intuitive. In fact, the idea was to have a log which is both easy to read by the analyst but also easy to parse and post-process.      
\\

@@

When tracing APIs, most of the concepts seen in Native API tracing also apply, but there are also significant differences, both in the steps that were carried out and in the Pin APIs which were employed, as it can be observed from the diagram in \textit{Figure 3.5}.

Now let us illustrate the actions being performed during API tracing by analyzing in detail each one of the three main blocks of \textit{Figure 3.5}.

@@

The first step taken towards APIs tracing is APIs instrumentation, i.e. placing analysis routines before and after the APIs' execution. To do this, a specific function was embedded in BluePill's image callback (the notion of image callback was previously introduced in Section 3.3.1). Such function, named \texttt{AddAnalysisRoutines}, performs the following actions for every DLL being loaded:

@@

Due to the high amount of APIs for which information was stored in the \texttt{PyREbox} database, it was decided to group API data by DLL. Specifically, for every DLL a corresponding \texttt{.cpp} file was created to store the API-related information under the form of \texttt{libcall\_arg\_info\_t} instances, one for each "known" API exported by that DLL. Furthermore, in order to make API data retrieval more efficient, it was chosen to employ a two-level hash map approach for this purpose. Given the DLL name (easily obtainable in the image callback via the use of \texttt{IMG\_Name}), the first level hash map is utilized to retrieve the DLL-specific second level hash map, which, in turn, maps the API name to the appropriate \texttt{struct} instance (\textit{Figure 3.6}). This is how API information retrieval is carried out in case \textit{PyREbox} database is used as API data source.
\\
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/HashMap.pdf}
\caption{\textit{PyREbox API Information Retrieval}}
\end{figure}

@@

In this section we will describe how APIs are analyzed before execution, i.e. the main steps that are carried out in the \texttt{PrintInfoBefore} functions, an overview of which is given in \textit{Figure 3.5}.
The workflow in this case is mostly the same as the one for NTAPIs' analysis, with one major difference, that is the introduction of a shadow stack used to store API entries in the thread local storage, which will be motivated later in the section.

@@

Lastly, let us detail the most important operations performed towards tracing API-related information after the API execution. These are carried out in the \texttt{PrintInfoAfter} function and can be observed from \textit{Figure 3.5}.

@@

Right before the execution of \texttt{KiUserCallbackDispatcher}, i.e. when the registered \texttt{CONTEXT\_CALLBACK\_NOTIFICATION} function is called, the stack looks as follows (\textit{Figure 3.7}).
\\
\begin{figure}[h]
\centering
\includegraphics[width=1\textwidth]{Figures/StackCallback.pdf}
\caption{\textit{Stack before \texttt{KiUserCallbackDispatcher} execution}}
\end{figure}

@@

\begin{enumerate}
\item The stack pointer value is retrieved. This is achieved through the use of the \texttt{PIN\_GetContextReg} API, that allows to get the value of a register in the context provided as input. Such context, in our specific case, is the application's register state after the context change, which, as previously mentioned, is passed as argument to the \texttt{CONTEXT\_CHANGE\_CALLBACK} notification function.
\item The callback number is obtained by adding 4 to the stack pointer and fetching the pointed value, in accordance to \textit{Figure 3.7}.
\item The \texttt{KernelCallbackTable} is retrieved from the PEB. 
\item The function pointer stored in the \texttt{KernelCallbackTable} entry indicated by the callback number is acquired.
\item It is checked if the address of the function falls within \texttt{User32.dll} memory address range. In fact, in \texttt{Win32} user mode callbacks are employed exclusively by \texttt{User32.dll} for windowing related aspects \cite{UCDispatcher}, meaning that the retrieved function address must belong to \texttt{User32.dll}'s memory address range. Such range is determined by retrieving an handle for \texttt{User32.dll} through \texttt{GetModuleHandle}, building an \texttt{IMG} object from it utilizing Pin's \texttt{IMG\_FindByAddress} and invoking the usual \texttt{IMG\_LowAddress} and \texttt{IMG\_HighAddress} functions. 
\item The name of the callback is determined by creating a \texttt{RTN} object from the function address via \texttt{RTN\_FindByAddress} and consequently invoking \texttt{RTN\_Name} on such object.
\item The callback occurrence and the callback name is recorded in the log.
\end{enumerate}

### 2018-07-02 introduction.tex
Chapter 2 Done
Blue Pill Added

Malicious software (or malware) is any software specifically designed to bring harm to a computer system. The problem posed by malwares is one which is becoming increasingly important, as new and more sophisticated malwares arise every day and the economical damage for organizations keeps worsening \cite{Cisco}. To face this threat, professionals are typically aided by a range of automatic tools capable of analysing and detecting malicious software.

Malware analysis can be carried out either statically or dynamically. Dynamic analysis encompasses techniques that execute a sample and observe the actions it actually performs, whereas in static analysis the sample is examined without running it. Such techniques have evolved over time to keep track with the increasing complexity and diversity of malwares. However, in recent years, the shift towards automation, caused by the need of dealing with a huge and ever-growing number of samples, together with the rising complexity of obfuscation mechanisms utilized by malwares, has strongly favoured dynamic analysis.

@@

A problem that all dynamic analysis techniques have to face, including function call monitoring by means of API hooking, is the widespread of evasive malwares. Such malwares check whether or not they are being executed in an adverse environment and conceal their harmful behaviours accordingly, like for example by carrying out an exit sequence \cite{BluePill}. Unfortunately, such anti-evasion mechanisms are frequently adopted my malicious samples. According to Symantec's Internet Security Report of March 2018, 18\% of new malware were virtual-machine-aware \cite{Symantec}. To make matters worse, the API hooking techniques presented in literature are easily detectable and are not coupled by any mechanism to hide their presence from evasive malwares.

The goal of this thesis is to present \textbf{BlueTracer}, a robust library and system call tracer for Windows applications, specialized in the monitoring of evasive malwares. BlueTracer is based on the Intel Pin \cite{Pin} dynamic binary instrumentation (DBI) framework and is able to counteract malwares' anti-evasion measures thanks to its integration with BluePill, a software toolkit built on top of a DBI layer which allows the simulation of the execution environment a particular malware was designed for and conceals any virtualization artifacts and setup details which might set off evasion \cite{BluePill}. BlueTracer is capable of tracing the input values, the output values and the return values of an extremely wide range of system calls (including Windows Native APIs) and API calls. Moreover, it also supports the tracing of Windows callbacks functions and Windows asynchronous procedure calls (APC). The tool was tested on a benign application aimed at assessing how good an anti-malware system is against evasion techniques and on actual evasive malwares, proving to be effective in both tracing the samples' activity and remaining undetected.

\paragraph{Structure of the Thesis.}
The remaining part of this thesis is structured as follows.
Chapter 1 describes the major \textit{API hooking} techniques present in literature, outlining their stengths and weaknesses, especially from a detection point of view.
Chapter 2 introduces the concept of Dynamic Binary Instrumentation (DBI) and presents Intel Pin, the framework used to develop BlueTracer.
Chapter 3 focuses on the implementation of the tool, on its structure and the design choices which were made during its development.
Chapter 4 illustrates the experimental results and assesses the tool's effectiveness.
Finally, we present concluding remarks in Chapter 5, together with possible future developments.

>>>>

Malicious software (or malware) is any software specifically designed to bring harm to a computer system. The problem posed by malware is one which is becoming increasingly important, as new and more sophisticated malware samples arise every day and the economical damage for organizations keeps worsening \cite{Cisco}. To face this threat, professionals are typically aided by a range of automatic tools capable of analysing and detecting malicious software.

Malware analysis can be carried out either statically or dynamically. Dynamic analysis encompasses techniques that execute a sample and observe the actions it actually performs, whereas in static analysis the sample is examined without running it. Such techniques have evolved over time to keep track with the increasing complexity and diversity of malware. However, in recent years, the shift towards automation, caused by the need of dealing with a huge and ever-growing number of samples, together with the rising complexity of obfuscation mechanisms utilized by malware, has strongly favoured dynamic analysis.

@@

A problem that all dynamic analysis techniques have to face, including function call monitoring by means of API hooking, is the widespread of evasive malware. The malware samples belonging to this category check whether or not they are being executed in an adverse environment and conceal their harmful behaviours accordingly, like for example by carrying out an exit sequence \cite{BluePill}. Unfortunately, such anti-evasion mechanisms are frequently adopted my malicious samples. According to Symantec's Internet Security Report of March 2018, 18\% of new malware were virtual-machine-aware \cite{Symantec}. To make matters worse, the API hooking techniques presented in literature are easily detectable and are not coupled by any mechanism to hide their presence from evasive malware.

\paragraph{Contributions.}
Throghout this thesis we present \textbf{BlueTracer}, a robust library and system call tracer for Windows applications, specialized in the monitoring of evasive malware. BlueTracer is based on the Intel Pin \cite{Pin} dynamic binary instrumentation (DBI) framework and is able to counteract malware's anti-evasion measures thanks to its integration with BluePill, a software toolkit built on top of a DBI layer which allows the simulation of the execution environment a particular malware sample was designed for and conceals any virtualization artifacts and setup details which might set off evasion \cite{BluePill}. BlueTracer is capable of tracing the input values, the output values and the return values of an extremely wide range of system calls (including Windows Native APIs) and API calls. Moreover, it also supports the tracing of Windows callbacks functions and Windows asynchronous procedure calls (APC). The tool was tested on a benign application aimed at assessing how good an anti-malware system is against evasion techniques and on actual evasive malware, proving to be effective in both tracing the samples' activity and remaining undetected.

\paragraph{Structure of the Thesis.}
The remaining part of this thesis is structured as follows.
Chapter 1 describes the major \textit{API hooking} techniques present in literature, outlining their stengths and weaknesses, especially from a detection point of view. It also introduces the concept of Dynamic Binary Instrumentation (DBI), which constitutes BlueTracer's core.
Chapter 3 focuses on the implementation of the tool, on its structure and the design choices which were made during its development.
Chapter 4 illustrates the experimental results and assesses the tool's effectiveness.
Finally, we present concluding remarks in Chapter 5, together with possible future developments.
